---
title: "Predicting weight lifting exercise quality"
author: "Shawn Estes"
date: "August 15, 2015"
output: html_document
---

### Overview

This is the course project for Coursera's Practical Machine Learning class.Our dataset includes user activity levels from accelerometers worn during weight lifting exercises. The goal is to predict which activity was performed based on a predefined dataset. 

### Data Sources

For detailed information on the dataset, see: http://groupware.les.inf.puc-rio.br/har in particular the section on Weight Lifting Exercises. 

The training dataset is available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The testing dataset (used only for the submitted portion of the assignment) is available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data is provided and used by the following: 

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013. 


Setup and Data Cleaning

R package Caret is required for running the code. We'll set a seed to aid reproducibility and then download files from the source if they are not already available locally.

Data cleaning is relatively simple. We need to be on the lookout for NA values that are not actually labeled in as NA. In this case, some values are received as "!#DIV/0" from the raw data. These values are useless and are overwritten as NA. 

Note: all transformations are applied directly to both training and testing splits. 

```{r}

library(caret)
library(randomForest)
set.seed(333)

trainingSetUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testingSetUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

if(!file.exists("data/pml-training.csv")) { 
  # Download the file 
  download.file(trainingSetUrl, destfile="./data/pml-training.csv", method="curl")
}

if(!file.exists("data/pml-testing.csv")) { 
  # Download the file 
  download.file(testingSetUrl, destfile="./data/pml-testing.csv", method="curl")
}


fullTrainingDataSet <- read.csv("./data/pml-training.csv", na.strings=c("NA","#DIV/0!",""), head=TRUE)
oos_test <- read.csv("./data/pml-testing.csv", na.strings=c("NA","#DIV/0!",""), head=TRUE)

# Partition the training data into 60/40 split 
inTrain <- createDataPartition(y=fullTrainingDataSet$classe, p=0.60, list=FALSE)
training <- fullTrainingDataSet[inTrain,]
testing <- fullTrainingDataSet[-inTrain,]
dim(training); dim(testing);

# Set any of the values of NA to 0 
training[is.na(training)] <- 0
#Remove near zero variables from training
if (length(nearZeroVar(training)) > 0) {
  training <- training[, -nearZeroVar(training)] 
}

# Duplicate transformations in testing
testing[is.na(testing)] <- 0
if (length(nearZeroVar(testing)) > 0) {
  testing <- testing[, -nearZeroVar(testing)] 
}

# And in the out of sample tests
oos_test[is.na(oos_test)] <- 0
if (length(nearZeroVar(oos_test)) > 0) {
  oos_test <- oos_test[, -nearZeroVar(oos_test)] 
}

# Verify dims between training and testing after cleaning
dim(training); dim(testing);
```

##Model fitting 

Random Forest is known for its accuracy, albeit not speed. With that in mind, this is a relatively small dataset so I've prioritized accuracy over speed. To greater reduce the speeds, I've chosen to subset the random forest to just the relevant columns. This also helps to prevent levels issues in new datasets.


```{r}

# Fit against a random forest
modFit <- randomForest(classe ~roll_belt+pitch_belt+magnet_dumbbell_y+magnet_dumbbell_z+pitch_forearm+accel_dumbbell_y+roll_arm+roll_forearm, data=training)
print(modFit)
```

### Estimate and testing of error rates

The random forest model has an out-of-bag (OOB) error of 1.88%. I'd suspect the same results when predicting against the test set on this model. Let's test it and show the confusion matrix results. 

```{r}
predictions <- predict(modFit, testing, type = "class")
# Show the results in a confusion matrix
confusionMatrix(predictions, testing$classe)
```

The accuracy rate of 98.1% matches closely with our OOB error estimates. Using the out of sample tests with the submission assignment, it succeeded in predicting all 20 of the problems on the first attempt. 

In conclusion, this random forest model should be able to predict the manner in which exercises were performed using this type of accelerometer data. 

```{r echo=FALSE}

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
test_predictions <- predict(modFit, newdata=oos_test, type = "class")
oos_test$classe <- test_predictions
submissions <- data.frame(problem_id = oos_test$problem_id, classe = test_predictions)
pml_write_files(submissions$classe)
```
